---
title: "Class_11"
author: "Brad Hunter (PID: A69038089)"
format: pdf
---

## Background

We saw last class that the PDB has 209,886 entries (Oct/Nov 2025). UniprotKB (i.e. protein sequence database) has 199,579,901 entries.

So the PDB has only `r round(209886/199579901 *100,2)`% coverage of the main sequence database.

Enter AlphaFold data base (AFDB) < http://alphafold.ebi.ac.uk > that attempts to provide computed models for all sequences in UniProt.

"AlphaFold DB provides open access to over 200 million protein structure predictions to accelerate scientific research."

## AlphaFold

AlphaFold has 3 main outputs

- The predicted coordinates (PDB file)
- A local quality score called pLDDT (one for each amino-acid)
- A second quality score called PAE, Predicted Aligned Error (for each pair of amino-acid)

We can run AlphaFold ourselves if we are not happy with AFDB (i.e. no coverage or poor model).



## Interpreting/analyzing AF results in R

```{r}
results_dir <- "Brads_HIVPR_dimer_23119/"
```

```{r}
# File names for all PDB models
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

```{r}
library(bio3d)
library(bio3dview)
# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

```{r}
pdbs
```
RMSD is a standard measure of structural distance between coordinate sets. We can use the `rmsd()` function to calculate the RMSD between all pairs models.


```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)
```

Drawing a heatmat
```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

Now lets plot the pLDDT values across all models.

```{r}
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")

rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

