---
title: "Class 7: Machine Learning 1"
author: "Brad Hunter PID:A69038089"
format: pdf
---

Today we will begin our exploration of some "classical" machine learing approaches. We will start with clustering:

First we'll make up some data to cluster with a known answer.

```{r}
hist(rnorm(1000))
```

```{r}
x <- c(rnorm(30, mean = -3), rnorm(30, mean = 3))
y <- rev(x)

z <- cbind(x,y)
```

Lets plot our data:

```{r}
plot(z)
```

The main function in "base" R for K-means clustering is called `kmeans()`.

```{r}
k <- kmeans(z, centers=2)
k
```

> Q. How big are the clusters (i.e their size)?

```{r}
k$size
```

> Q. What clusters do my data points reside in?

```{r}
k$cluster
```

> Q. Make a plot of our data colored by cluster assignment - i.e. Make a result figure...

```{r}
plot(z, col = k$cluster)
points(k$centers, col ="blue", pch=15)
```

>Q. Cluster with k-means into 4 clusters and plot your results as above

```{r}
k4 <- kmeans(z, centers=4)
plot(z, col = k4$cluster)
points(k4$centers, col ="blue", pch=15)
```


>Q. Run kmeans with centers (i.e. values of k) equal 1 to 6, using a for loop

```{r}
ans <- NULL
for(i in 1:6){
  ans <- c(ans, kmeans(z, centers = i)$tot.withinss)
}
ans
```

Plot this "scree-plot"

```{r}
plot(ans, typ="b")
```

## Hierarchical Clustering

The main function in "base" R for this is called `hclust()`.

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

```{r}
plot(hc)
abline(h=7, col="red")
```

To obtain clusters from our `hclust` object **hc** we "cut" the tree to yield different sub branches. For this we use the `cutree()` function.

```{r}
grps <- cutree(hc, h=7)
grps
```

Results Figure

```{r}
plot(z, col=grps)
```

## PCA's!!!!! (Principal Component Analysis)

```{r}
url <- "https://tinyurl.com/UK-foods"
data <- read.csv(url, row.names = 1)
```


>Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(data)
```

>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the second method (read.csv(url, row.names=1)) so that my row names are not lost with no danger to losing data (like calling x <- x[,-1] multiple times)

### Spotting major differences and trends

```{r}
barplot(as.matrix(data), beside=T, col=rainbow(nrow(data)))
```

>Q3: Changing what optional argument in the above barplot() function results in the following plot?

Changing the besides arugment to false will result in the following plot

```{r}
barplot(as.matrix(data), beside=F, col=rainbow(nrow(data)))
```

>Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

If a given point lies on the diagonal, it is the same (or very similar) for both datasets (countries) being compared.

```{r}
pairs(data, col=rainbow(10), pch=16)
```

>Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

It's hard to tell, it seems like N. Ireland consumes more dark blue food group. Wales and England look very similar.

## PCA to the rescue 

The main function in "base" R for PCA is called `prcomp()`.

But first we must transpose our dataset so the foods (dependent var) are in the columns and the countries (independent var) are in the rows using the `t()` function.

```{r}
pca <- prcomp(t(data))
summary(pca)
```
Our result object is called `pca` and it has a `$x` componenent that we will look at first

```{r}
pca$x
```
Making a ordination plot

```{r}
library(ggplot2)

ggplot(pca$x,aes(PC1,PC2, label=rownames(pca$x))) +
  geom_point() +
  geom_text()
```

Another major result out of PCA is the so-called "variable loadings" or `$rotation` that tells us how the origional variables (foods) contribute to PCs (i.e. our new axis).

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```


>Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

