---
title: "Genomics Homework"
author: "Brad Hunter (PID: A69038089)"
format: pdf
---

## Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

```{r, fig.show='hide'}
data <- read.table("rs8067378_ENSG00000172057.6.txt")

x <- boxplot(exp ~ geno, data = data)

stats_table <- data.frame(
  Genotype = x$names,
  Sample_Size = x$n,
  Median = round(x$stats[3, ], 2)
) 
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
flextable::flextable(stats_table)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(data, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot() +
  guides(fill = "none", color = "none") +
  labs(
    title = "ORMDL3 Expression by Genotype",
    x = "Genotype",
    y = "Expression Level") +
  theme_minimal()
```

It does seem that SNP effects the expression of ORMDL3, the A|A genotype has a higher median experession level compared to the G|G genotype.