---
title: "Lab_10_Halloween_Project"
author: "Brad Hunter (PID: A69038089)"
format: pdf
---

## Happy Halloween!!!

Let's load libraries up top for sanity.
```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library("skimr")
library(ggrepel)
library(corrplot)
```

Let's load and look at our data.
```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names = 1)
flextable::flextable(head(candy))
```

> Q1. How many different candy types are in this dataset?

There are `r ncol(candy)` different candy types (categories) and `r nrow(candy)` different types of candy.

> Q2. How many fruity candy types are in the dataset?

There are `r sum(candy$fruity)` fruity candy types.

## What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Almond Joy and it's win percentage is `r round(candy["Almond Joy",]$winpercent,2)`%.
```{r}
candy["Almond Joy",]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

Kit Kat's win percentage is `r round(candy["Kit Kat",]$winpercent,2)`%.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

The win percentage for Tootsie Roll Snack Bars is `r round(candy["Tootsie Roll Snack Bars",]$winpercent,2)`%.

Let's look at our data using skimr

```{r}
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, winpercent looks to be on a 100x scale than the other columns.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero represents a candy without any chocolate, and a 1 represents a candy with chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 20)
```

> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```

No, it does not look symmetrical.

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

Below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.win.ave <- candy |> filter(chocolate==1) |> pull(winpercent) |> mean()
fruit.win.ave <- candy |> filter(fruity==1) |> pull(winpercent) |> mean()
```

Yes, the chocolate win mean is `r choc.win.ave` and the fruity win mean is `r fruit.win.ave`.

> Q12. Is this difference statistically significant?

```{r}
t.test((candy |> filter(chocolate==1) |> pull(winpercent)), candy |> filter(fruity==1) |> pull(winpercent))
```
Yes, the difference is significantly different.

> Q13. What are the five least liked candy types in this set?

```{r}
candy |> arrange(winpercent) |> head(5) |> rownames()
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |> arrange(winpercent) |> tail(5) |> rownames()
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

Let's add some color to the graph!

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "orange"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 

```

> Q17. What is the worst ranked chocolate candy?

The worst ranked choclolate is `r candy |> filter(chocolate==1) |> arrange(winpercent) |> head(1) |> rownames()`.


> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is `r candy |> filter(fruity==1) |> arrange(winpercent) |> tail(1) |> rownames()`.

## Taking a look at pricepercent

```{r, warning=FALSE}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy |> arrange(pricepercent) |> tail(5) |> arrange(winpercent) |> rownames()
```

And of those, Nik L Nip is the least popular.

## Correlation

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate are anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are most correlated, followed closely by chocolate and bar.

## Principal Component Analysis

The main function in base R for this is `prcomp()` and we want to set `scale=TRUE` here:

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
Let's plot our main PCA scores

```{r}
ggplot(pca$x) +
  aes(PC1,PC2, label=rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col=my_cols) +
  labs(title="Brad's Candy Analysis")
```

Don't forget about your variable "loadings" - how the origional variables contribute to your new PCs...

```{r}
ggplot(pca$rotation) +
  aes(PC1,rownames(pca$rotation)) +
  geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Chocolate, bar, winpercent, pricepercent are all strongly positive in PC1 and yes this makes sense, all of these variables are associated with each other.